---
title: "Lab 05 - Data Wrangling"
output: 
  github_document: default
  tufte::tufte_html:
    css: style.css
link-citations: yes
always_allow_html: true
---

# Learning goals

- Use the `merge()` function to join two datasets.
- Deal with missings and impute data.
- Identify relevant observations using `quantile()`.
- Practice your GitHub skills.

# Lab description

For this lab we will be dealing with the meteorological dataset `met`. In this case, we will use `data.table` to answer some questions regarding the `met` dataset, while at the same time practice your Git+GitHub skills for this project.

This markdown document should be rendered using `github_document` document.

# Part 1: Setup a Git project and the GitHub repository

1. Go to wherever you are planning to store the data on your computer, and create a folder for this project

2. In that folder, save [this template](https://github.com/JSC370/JSC370-2024/blob/main/labs/lab05/lab05-wrangling-gam.Rmd) as "README.Rmd". This will be the markdown file where all the magic will happen.

3. Go to your GitHub account and create a new repository of the same name that your local folder has, e.g., "JSC370-labs".

4. Initialize the Git project, add the "README.Rmd" file, and make your first commit.

5. Add the repo you just created on GitHub.com to the list of remotes, and push your commit to origin while setting the upstream.

Most of the steps can be done using command line:

```sh
# Step 1
cd ~/Documents
mkdir JSC370-labs
cd JSC370-labs

# Step 2
wget https://raw.githubusercontent.com/JSC370/JSC370-2024/main/labs/lab05/lab05-wrangling-gam.Rmd
mv lab05-wrangling-gam.Rmd README.Rmd
# if wget is not available,
curl https://raw.githubusercontent.com/JSC370/JSC370-2024/main/labs/lab05/lab05-wrangling-gam.Rmd --output README.Rmd

# Step 3
# Happens on github

# Step 4
git init
git add README.Rmd
git commit -m "First commit"

# Step 5
git remote add origin git@github.com:[username]/JSC370-labs
git push -u origin master
```

You can also complete the steps in R (replace with your paths/username when needed)

```r
# Step 1
setwd("~/Documents")
dir.create("JSC370-labs")
setwd("JSC370-labs")

# Step 2
download.file(
  "https://raw.githubusercontent.com/JSC370/JSC370-2024/main/labs/lab05/lab05-wrangling-gam.Rmd",
  destfile = "README.Rmd"
  )

# Step 3: Happens on Github

# Step 4
system("git init && git add README.Rmd")
system('git commit -m "First commit"')

# Step 5
system("git remote add origin git@github.com:[username]/JSC370-labs")
system("git push -u origin master")
```

Once you are done setting up the project, you can now start working with the MET data.

## Setup in R

1. Load the `data.table` (and the `dtplyr` and `dplyr` packages), `mgcv`, `ggplot2`, `leaflet`, `kableExtra`.

```{r packages, warnings=FALSE}
library(data.table)
library(dtplyr)
library(dplyr)
library(mgcv)
library(ggplot2)
library(leaflet)
library(kableExtra)
library(tidyr)
library(knitr)
library(styler)
library(webshot)
```


```{r, eval=TRUE}
fn <- "https://raw.githubusercontent.com/JSC370/JSC370-2024/main/data/met_all_2023.gz"
if (!file.exists("met_all_2023.gz"))
  download.file(fn, destfile = "met_all_2023.gz")
met <- data.table::fread("met_all_2023.gz")
```

2. Load the met data from https://github.com/JSC370/JSC370-2024/main/data/met_all_2023.gz or  (Use https://raw.githubusercontent.com/JSC370/JSC370-2024/main/data/met_all_2023.gz to download programmatically), and also the station data. For the latter, you can use the code we used during lecture to pre-process the stations data:

```{r stations-data, eval = TRUE}
# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE, LAT, LON)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

# Read in the met data and fix lat, lon, temp
met$lat <- met$lat/1000
met$lon <- met$lon/1000
met$wind.sp <- met$wind.sp/10
met$temp <- met$temp/10
met$dew.point <- met$dew.point/10
met$atm.press <- met$atm.press/10
```

3. Merge the data as we did during the lecture. Use the `merge()` code and you can also try the tidy way with `left_join()`

```{r}
mer <- merge(met, stations, 
             by.x = "USAFID", by.y="USAF",
             all.x = TRUE, all.y = FALSE)
```

## Question 1: Identifying Representative Stations

Across all weather stations, which stations have the median values of temperature, wind speed, and atmospheric pressure? Using the `quantile()` function, identify these three stations. Do they coincide?


```{r}
# Calculate the median values for temperature, wind speed, and atmospheric pressure
national_median_temp <- quantile(mer$temp, probs = 0.5, na.rm = TRUE)
national_median_wind_speed <- quantile(mer$wind.sp, probs = 0.5, na.rm = TRUE)
national_median_atm_pressure <- quantile(mer$atm.press, probs = 0.5, na.rm = TRUE)
```

Next identify the stations have these median values.

```{r}
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
mer_by_stations <- mer %>% 
  # filter(!is.na(temp) & !is.na(wind.sp) & !is.na(atm.press)) %>%
  group_by(USAFID) %>% 
  summarise(
    med_temp = median(temp, na.rm = TRUE),
    med_wind_speed = median(wind.sp,na.rm = TRUE),
    med_atm_press = median(atm.press,na.rm = TRUE),
    state = get_mode(STATE),
    lat = median(lat),
    lon = median(lon),
    elev = median(elev)
            )
nrow(mer_by_stations)
```


```{r}
# Find stations where these median values occur
stations_median_temp <- unique(mer_by_stations$USAFID[which(mer_by_stations$med_temp == national_median_temp)])
stations_median_wind_speed <- unique(mer_by_stations$USAFID[which(mer_by_stations$med_wind_speed == national_median_wind_speed)])
stations_median_atm_pressure <- unique(mer_by_stations$USAFID[which(mer_by_stations$med_atm_press == national_median_atm_pressure)])
# Output the stations with median values
cat("Stations with median temperature:", stations_median_temp, "\n")
cat("Stations with median wind speed:", stations_median_wind_speed, "\n")
cat("Stations with median atmospheric pressure:", stations_median_atm_pressure, "\n")

common_stations <- Reduce(intersect, list(stations_median_temp, stations_median_wind_speed, stations_median_atm_pressure))
cat("Stations with national median temperature, wind speed, and atmospheric pressure:", common_stations)
```

Knit the document, commit your changes, and save it on GitHub. Don't forget to add `README.md` to the tree, the first time you render it.

## Question 2: Identifying Representative Stations per State

Now let's find the weather stations by state with closest temperature and wind speed based on the euclidean distance from these medians. 
```{r}
temp_windsp_dist <- function(station_temp, state_temp, station_wsp, state_wsp) {
  sqrt((state_temp - station_temp)^2+(state_wsp - station_wsp)^2)
}

mer_by_states <- mer %>% 
  group_by(STATE) %>% 
  summarise(
    state_med_temp = median(temp, na.rm = TRUE),
    state_med_wind_speed = median(wind.sp,na.rm = TRUE),
    state_med_atm_press = median(atm.press,na.rm = TRUE),
    state_med_lat = median(lat),
    state_med_lon = median(lon)
  )
mer_by_states
```



```{r}
mere_by_stations <- mer_by_stations %>%
  merge(mer_by_states) %>% 
  filter(state == STATE) %>% 
  mutate(dist = temp_windsp_dist(state_temp = state_med_temp,
                                 station_temp = med_temp,
                                 state_wsp = state_med_wind_speed,
                                 station_wsp = med_wind_speed))
  


# Group by state and find closest station
closest_stations <- mere_by_stations %>%
  group_by(state) %>%
  top_n(-1, dist) %>%
  ungroup()

# Select only necessary columns
closest_stations <- closest_stations %>%
  select(USAFID, med_temp, med_wind_speed, state, lat, lon, elev, dist, ) %>% 
  unique()

closest_stations
```

Knit the doc and save it on GitHub.

## Question 3: In the Geographic Center?

For each state, identify which station is closest to the geographic mid-point (median) of the state. Combining these with the stations you identified in the previous question, use `leaflet()` to visualize all ~100 points in the same figure, applying different colors for the geographic median and the temperature and wind speed median.

```{r}
geo_distance <- function(lat1, lon1, lat2, lon2) {
  distance <- sqrt((lat2 - lat1)^2 + (lon2 - lon1)^2)
  return(distance)
}

mere_by_stations <- mere_by_stations %>%
  mutate(geo_dist = geo_distance(lat, lon, state_med_lat, state_med_lon))

# Find the closest station to the geographic midpoint of each state
geo_closest_stations_by_state <- mere_by_stations %>%
  group_by(state) %>%
  arrange(geo_dist) %>%
  slice(1)

# Create the leaflet map
map <- leaflet(geo_closest_stations_by_state) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    color = ~"purple",
    radius = 5,
    #popup = ~paste("USAFID: ", USAFID, "<br>",
    #               "Median Temperature: ", med_temp, "<br>",
    #               "Median Wind Speed: ", med_wind_speed, "<br>",
    #               "State: ", state, "<br>",
    #               "Latitude: ", lat, "<br>",
    #               "Longitude: ", lon, "<br>",
    #               "Elevation: ", elev, "<br>",
    #               "Geo distance: ", geo_dist, "<br>") 
  ) %>% 
  addCircleMarkers(
    data = closest_stations,
    lng = ~lon,
    lat = ~lat,
    color = ~"orange",
    radius = 2
  ) %>% 
  addLegend(colors = c("purple","orange"), labels = c("Geographic closest", "Temperature and wind speed closest"))
map
```

Knit the doc and save it on GitHub.

## Question 4: Summary Table with `kableExtra`

Generate a summary table using `kable` where the rows are each state and the columns represent average temperature broken down by low, median, and high elevation stations. 

Use the following breakdown for elevation:

- Low: elev < 93
- Mid: elev >= 93 and elev < 401
- High: elev >= 401

```{r}
mer <- mer %>%
  mutate(elev_category = case_when(
    elev < 93 ~ "Low",
    elev >= 93 & elev < 401 ~ "Mid",
    elev >= 401 ~ "High"
  ))

# Calculate average temperature for each elevation category within each state
summary_table <- mer %>%
  group_by(STATE, elev_category) %>%
  summarise(avg_temp = mean(temp, na.rm = TRUE),.groups = "drop")

# Reshape the data to have elevation categories as columns
summary_table <- spread(summary_table, key = elev_category, value = avg_temp)

# Add row names for better readability
rownames(summary_table) <- summary_table$STATE

# Print the summary table using kable
kable(summary_table, caption = "average temperature by low, median, and high elevation stations", format = "pipe") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```



Knit the document, commit your changes, and push them to GitHub. 


## Question 5: Advanced Regression

Let's practice running regression models with smooth functions on X. We need the `mgcv` package and `gam()` function to do this.

- using your data with the median values per station, first create a lazy table. Filter out values of atmospheric pressure outside of the range 1000 to 1020. Examine the association between temperature (y) and atmospheric pressure (x). Create a scatterplot of the two variables using ggplot2. Add both a linear regression line and a smooth line.

- fit both a linear model and a spline model (use `gam()` with a cubic regression spline on atmospheric pressure). Summarize and plot the results from the models and interpret which model is the best fit and why.

```{r}

lazy_table <- mere_by_stations %>%
  filter(med_atm_press >= 1000 & med_atm_press <= 1020)

scatterplot <- ggplot(lazy_table, aes(x = med_atm_press, y = med_temp)) +
  geom_point() +  
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Linear regression line
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cr"), se = FALSE, color = "red") +  # Smooth line with cubic regression spline
  labs(x = "Atmospheric Pressure", y = "Temperature") +  # Labels
  theme_minimal()

print(scatterplot)

linear_model <- lm(med_temp ~ med_atm_press, data = lazy_table)

spline_model <- gam(med_temp ~ s(med_atm_press, bs = "cr"), data = lazy_table)

summary(linear_model)

summary(spline_model)

plot(linear_model)
plot(spline_model)
```

- the GAM (gam) model appears to be a better fit for the data 
  compared to the linear model (lm) since The GAM has a higher 
  adjusted R-squared value (0.297) compared to the adjusted 
  R-squared value of the linear model (0.2638), 
  indicating that the GAM explains more variance in the response variable.
- The GAM also has a lower residual standard error, 
  suggesting better fit to the data.
- This might because the smooth term in the GAM captures 
  the non-linear relationship between med_temp and med_atm_press.



## Deliverables

- .Rmd file (this file)

- link to the .md file (with all outputs) in your GitHub repository