---
title: "Lab 9 - HPC"
output: 
  # pdf_document: default
  # html_document: default
  tufte::tufte_html:
    css: style.css
link-citations: yes
---

# Learning goals

In this lab, you are expected to learn/put in practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the parallel package.
- Use Rscript to submit jobs.

```{r echo=FALSE}
# install any missing packages
# install.packages("microbenchmark")
library(microbenchmark)
library(ggplot2)
library(parallel)
```

## Problem 1: Think

Give yourself a few minutes to think about what you learned about parallelization. List three
examples of problems that you believe may be solved using parallel computing,
and check for packages on the HPC CRAN task view that may be related to it.

Three examples of problems with parallel computing:

Generating samples using Monte Carlo methods.

Computing forward pass in deep learning models.

Feature extraction in image processing.

Function that may be related to it:

parallel: Provides functions for parallel execution of R code on multiple cores or clusters of machines.

foreach: Allows for parallel execution of loops using different backends, such as parallel, doParallel, doSNOW, etc., enabling users to easily parallelize tasks.

doParallel: Provides a parallel backend for the foreach package, allowing parallel execution of loops on multicore machines.

## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without using
`parallel`:

1. This function generates a `n x k` dataset with all its entries having a Poisson distribution with mean `lambda`.

```{r p2-fun1}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), ncol = k)
}

# Benchmarking
microbenchmark::microbenchmark(
  fun1(100),
  fun1alt(100)
)
```

How much faster?

_unvectorized code mean: 312.37244; vectorized code mean: 35.77742, so around 277 microseconds faster._


2.  Find the column max (hint: Checkout the function `max.col()`).

(There isn't a built-in function in R named max.row that directly finds the row index of the maximum value in each column of a matrix. However, you can achieve this functionality using various approaches. One simple approach is to transpose the matrix, then use max.col to find the column index of the maximum value in each row of the transposed matrix. This effectively finds the row index of the maximum value in each column of the original matrix.)
(To summarize, max.col() returns the column index of the max in each row of the matrix.)

```{r p2-fun2}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  x[cbind(max.col(t(x)), 1:ncol(x))]
}

# Benchmarking
mb <- microbenchmark::microbenchmark(
  fun2(x),
  fun2alt(x)
)

# homework
autoplot(mb)
```


## Problem 3: Parallelize everything

We will now turn our attention to non-parametric 
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.

The main assumption is that we can approximate many experiments by resampling
observations from our original dataset, which reflects the population. 

This function implements the non-parametric bootstrap:

```{r p3-boot-fun}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
 
  # Making the cluster using `ncpus`
  # STEP 1: GOES HERE
  cl <- makePSOCKcluster(ncpus)
  # STEP 2: GOES HERE
  clusterExport(cl, varlist = c("idx", "dat", "stat"), envir = environment())
  
  # STEP 3: THIS FUNCTION NEEDS TO BE REPLACED WITH parLapply
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  # STEP 4: GOES HERE
  stopCluster(cl)
  
  ans
  
}
```

1. Use the previous code, and make it work with `parallel`. Here is just an example
for you to try:

```{r p3-test-boot}
# Bootstrap of a linear regression model
my_stat <- function(d) {
  coef(lm(y ~ x, data = d)) 
  }

# DATA SIM
set.seed(1)
n <- 500 
R <- 1e4
x <- cbind(rnorm(n)) 
y <- x*5 + rnorm(n)

# Check if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(data.frame(x, y), my_stat, R, ncpus = 8)

ans0
t(apply(ans1, 2, quantile, probs = c(0.025, 0.975)))
```

2. Check whether your version actually goes faster than the non-parallel version:

```{r benchmark-problem3}
# your code here
parallel::detectCores()
system.time(my_boot(data.frame(x, y), my_stat, R, ncpus = 1))
system.time(my_boot(data.frame(x, y), my_stat, R, ncpus = 8))
```

_Using more number of CPU cores takes more time in user and system but elapsed time in which we run the code is faster._

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command
in your terminal:

```bash
Rscript --vanilla -e 'rmarkdown::render("[full-path-to-your-Rmd-file.Rmd]")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full path to
your Rmd file... :).


